Move Schedules: Fast persistence computations in sparse dynamic settings \thanks{\textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.}

### 2021-09-03 move_schedules.tex
Updates from Overleaf

% REQUIRED - "Abstracts must be able to stand alone and so cannot contain citations to the paper's references, equations, etc. 
\begin{abstract}
	The standard procedure for computing the persistent homology of a filtered simplicial complex is the matrix reduction algorithm. Its output is a particular decomposition of the total boundary matrix, from which the persistence diagrams and generating cycles can be derived. 
	Persistence diagrams are known to vary continuously with respect to their input; this motivates the algorithmic study of persistence computations for time-varying filtered complexes. Computationally, simulating persistence dynamically can be reduced to maintaining a valid decomposition under adjacent transpositions in the filtration order. In practice, the quadratic scaling in the number of  transpositions often makes this maintenance procedure slower than simply computing the decomposition from scratch, effectively limiting the application of dynamic persistence to relatively small data sets. In this work, we propose a coarser strategy for maintaining the decomposition over a discrete 1-parameter family of filtrations. Our first result is an analysis of a simple linear-time strategy which reduces the number of column operations needed to simulate persistence across a fixed homotopy by at most a factor of 2. We then show a modification of this technique which maintains only a sublinear number of valid states, as opposed to a quadratic number of states, and we provide tight lower bounds for this technique.
	Finally, we provide empirical results suggesting that the decrease in operations needed to compute diagrams across a family of filtrations is proportional to the difference between the expected quadratic number of states, and the proposed sublinear coarsening.
\end{abstract}

@@

\section{Introduction} 
\subsection{Overview}\label{sec:overview} 
The starting point to the work described in this paper is the stability results established in~\cite{cohen2007stability, cohen2006vines}. 
Indeed, given a triangulable topological space equipped with a sufficiently tame continuous function, persistent homology captures the changes in topology across the sublevel sets of the space, and encodes them in a persistence diagram. The stability of persistence contends that small changes in the underlying function produce small changes in the corresponding diagrams:  if the function changes continuously, so too will the points on the persistence diagram. 
This motivates the study and application of persistence to time-varying or dynamic settings, like that of dynamic metric spaces~\cite{kim2020spatiotemporal}. 
As persistence-related computations tend to exhibit high algorithmic complexity (e.g. essentially cubic\footnote{For finite fields, it is known that the persistence computation reduces to the PLU factorization problem, which takes $O(m^\omega)$ where $\omega \approx 2.373$ is the matrix multiplication constant.} in the size of the underlying filtration~\cite{morozov2005persistence}), their adoption to dynamic settings poses a challenging computational problem.
With the current state of the art tools, there is no recourse to the problem of computing persistence from a time-varying filtration containing millions of simplices at thousands of snapshots in time.
An efficient means of updating the persistence of a [time-varying] filtration has far-reaching consequences: methods that vectorize persistence diagrams for machine learning purposes, like adaptive template functions~\cite{polanco2019adaptive} and persistence images~\cite{adams2017persistence}, or summary statistics like $\alpha$-smoothed Betti curves~\cite{ulmer2019topological}, all immediately become computationally viable tools in dynamic  settings. 
%This analogy provides a strong interpretation at gaining insight on the subtle topological changes that may occur in studying a continuous process.

Cohen-Steiner et al. refer to a continuous 1-parameter family of persistence diagrams as a \emph{vineyard}, and they give in \cite{cohen2006vines} an efficient algorithm for their computation given a time-varying filtration. 
The vineyards approach can be interpreted as an extension of the \emph{matrix reduction} algorithm~\cite{zomorodian2005computing}, which computes the persistence diagram of a fixed filtration $K$ with $m$ simplices in $O(m^3)$ time.
The main step in the reduction algorithm is a specific decomposition $R = D V$ of the boundary matrix $D$ of $K$. 
Since $V$ is always full rank (and upper-triangular), an alternative is to compute a $D = RU$ decomposition, where $U = V^{-1}$. 
The vineyards algorithm transforms a time-varying filtration into an ordered set of adjacent transpositions to be applied to the rows and columns of the decomposition $D = RU$, each of which take at most $O(m)$ time to execute. If one is interested in understanding how the persistent homology of a continuous function changes over time, then this algorithm is sufficient, for  homological critical points can only occur when the filtration order changes. 
Indeed, this algorithm is also efficient  asymptotically: if there are $d$ time-points where this filtration order changes, then the vineyards algorithm takes $O(m^3 + md)$ time: one $O(m^3)$-time reduction upfront for the filtration at time $t_0$ followed by one $O(m)$ operation to update the decomposition for the remaining time points $(t_1, t_2, \dots, t_d)$. When $d >> m$, this $O(md)$ approach is far more efficient than the $O(dm^3)$  strategy of computing the diagrams independently at every time point, what we call the ``naive'' approach.

Despite its efficient in theory, vineyards is often not the method of choice in practice in dynamic settings. 
While there is an increasingly rich ecosystem of software packages offering variations of the standard reduction algorithm (e.g. Ripser, PHAT, Dionysus, etc... see~\cite{otter2017roadmap} for an overview), implementations of the vineyards algorithm are relatively uncommon.\footnote{Dionysus 1 has an implementation of vineyards, but the algorithm was never ported to the 2nd version.} The reason for this disparity is perhaps explained by Lesnick and Wright~\cite{lesnick2015interactive}: ``While an update to an $RU$ decomposition involving few transpositions is very fast in practice... many transpositions can be quite slow... it is sometimes much faster to simply recompute the $RU$-decomposition from scratch using the standard persistence algorithm.'' Indeed, whether performing the reductions naively or updating the decomposition dynamically using vineyards, they observed that obtaining the decomposition is the most computationally demanding aspect of their 2-D persistence algorithm.

This work seeks to further understand and remedy this discrepancy: building on the work presented in~\cite{busaryev2010tracking}, we introduce a coarser approach to the vineyards algorithm.
While the vineyards algorithm is efficient at constructing a \emph{continuous} time family of diagrams, it is not necessarily efficient when the time parameter is coarsely discretized.
Our methodology is based on the observation that practitioners often don't need (or want!) all $d$ persistence diagrams generated by a continuous 1-parameter of filtrations; usually just $n << d$ of them are sufficient.   
% in many time varying settings, a valid decomposition is often only sought after at a relatively sparse number of time points relative to the size of the (finite) set needed to generated by the family. 
By exploiting the ``donor'' concept introduced in in~\cite{busaryev2010tracking}, we are able to make a tradeoff in the number of times the decomposition is restored to a valid state, dramatically reducing the total number of column operations needed to apply an arbitrary permutation to the filtration. This tradeoff, paired with a fast greedy heuristic explained in section~\ref{sec:proxy_objective}, yields an algorithm that can update a $R = DV$ decomposition more efficiently than vineyards in sparse contexts, making dynamic persistence more computationally tractable for a wider class of use-cases.

@@

 \subsection{A Motivating Example and Contributions}\label{sec:motivation} 
As a motivating example, we describe a simple experiment illustrating why the vineyards algorithm does not always yield an efficient method in many time-varying applications. 
Consider a series of grayscale images (i.e. a video) depicting a fixed-width annulus expanding about the center of a $9 \times 9$ grid, where scale of the annulus is parameterized by a single parameter $\tau$. Thus, the data set is a 1-parameter family of images. Each image contains the same number of pixels whose intensities---these are determined by the parameter---vary with time. For each image, we build a filtered simplicial complex using the \emph{Freudenthal} triangulation, with the ordering of inclusions obtained by lower stars.  
 \begin{figure*}[t]
 	\centering
 	\includegraphics[width=0.36\textwidth, height = 1.45in]{circle_vineyards.png}
 	\includegraphics[width=0.63\textwidth, height = 1.45in]{circle_complex.png}
 	\caption{On the right, the grayscale image data set and corresponding lower star filtrations. On the left, the cumulative number of column operations required to compute persistence for this time-varying filtration. The persistent Betti numbers for each filtration from left to right are: $\beta_1 = (1,0,0), \beta_2 = (1,1,0), \beta_3 = (1,1,0), \beta_4 = (1,1,0), \beta_5 = (4,0,0)$. Observe computing $10$ diagrams independently at evenly spaced time points (the green line) both captures these major topological changes and is the most computationally efficient approach shown.}
 	\label{fig:vineyards}
 \end{figure*}
Snapshots of the image data and its corresponding time-varying filtration is depicted in Figure~\ref{fig:vineyards} (top right and bottom right, respectively), which we use as a baseline for comparing the standard algorithm $\texttt{pHcol}$ with vineyards. Note the size of the filtration is fixed--no simplices are ever added or removed over time. 
There are two events which dramatically alter the persistence diagrams extracted from these filtrations: the first occurs when the central connected component splits to form a cycle, and the second occurs when the annulus approaches the outer dimensions of the grid, splitting into four components. As a consequence, in this example, only a few persistence diagrams are needed to capture the major topological changes over time.

  Consider the situation where one is concerned with capturing these dramatic changes in homology.
  On the left side of Figure~\ref{fig:vineyards}, we compare the cumulative cost (in total number of column operations) of various approaches which one can use to simulate persistence dynamically. 
  Since it is unknown a priori at what points in time the persistent homology will change in application settings, one solution is to discretize the time domain at $n$ evenly spaced points and compute the persistence diagram at each of these time points independently. An alternative approach is to construct a homotopy between two time-points, and then decompose this homotopy into adjacent transpositions--the vineyards approach. The former is often used in practice, but it is only an approximation; the latter is guaranteed to capture all homological changes in persistence that occur in simulating the 1-parameter family. The graph in Figure~\ref{fig:vineyards} depicts two different size approximations (purple, $n=100$ and green, $n=10$), wherein the reduction algorithm (\texttt{pHcol}) was applied at $n$ time points, and two exact strategies (black and blue), which compute all $\approx 7 \times 10^{4}$ diagrams given by the homotopy. As one can see from the figure, if one needs to simulate the homotopy exactly, the vineyards approach is indeed far more efficient than naively applying the reduction algorithm independently at all time points.
  However, when the discretization of the time domain is sparse enough, the naive approach actually performs less column operations than the vineyards strategy, while still capturing the main events.

@@

The original purpose of the vineyards algorithm, as described in~\cite{cohen2006vines}, was to compute a continuous 1-parameter family of persistence diagrams over a time-varying filtration, and to detect homological critical events during the construction.

@@

\begin{corollary}\label{cor:expectation}
	If $K_0, K_1$ are random distinct filtrations each of size $m$, then $d$ is upper bounded by $O(m - \sqrt{m})$ in expectation, with probability $1$ as $m \to \infty$.
\end{corollary}
\begin{proof}
	The proof of this result reduces to showing the average length of the LIS for random permutations. Let $L(p) \in [1,m]$ denote the maximal length of a increasing subsequence of $p \in S_m$. 
	The essential quantity to show the expected length of $L(p)$ over all permutations: 
	$$ \ell_m = \mathbb{E} \, L(p) = \frac{1}{m!} \sum\limits_{p \in S_m} L(p)$$
	The history of the mathematics of estimating this quantity dates back at least 50 years, and is sometimes called the \emph{Ulam-Hammersley} problem. The seminal work by Baik et al.~\cite{baik1999distribution} established that as $m \to \infty$:
	$$ \displaystyle \ell_m = 2 \sqrt{m} + c m^{1/6} + o(m^{1/6}) $$
where $c = -1.77108...$. As a result, we have: 
$$ \frac{\ell_m}{\sqrt{m}} \to 2\rlap{\quad \text{as } $m \to \infty$} $$	 
Thus, if $p \in S_m$ denotes a uniformly random permutation in $S_m$, then $L(p)/\sqrt{m} \to 2$ in probability as $m \to \infty$. 
Since the schedule size is defined as $d = m - \lvert \mathrm{LCS}(K_0, K_1) \rvert$, the claimed bound follows.
\end{proof}
\noindent The upper bound from Corollary~\ref{cor:expectation} captures the size of $S^\ast$ under worst-case conditions. It is worth noting that this bound applies to pairs of uniformly sampled permutations, as opposed to uniformly sampled distinct filtrations---intuitively, simplicial filtrations have more structure due to the subcomplex relation that must be respected in the filtration. Thus, while $O(m - \sqrt{m})$ in expectation is an improvement over $O(m)$, it is still not tight. However, Boissonnat~\cite{boissonnat2018efficient} has shown that for any $d$-dimensional simplicial complex $K$ with $m$ simplices, the number of distinct filtrations built from $K$ is \emph{at least} $\floor{\frac{t+1}{d+1}}^m$, where $t$ is the number of distinct filtration values associated with the complex $K$. In the case where $t \sim O(m)$ and $d << m$ fixed, this lower bound grows similarly to $m!$---thus, $O(m - \sqrt{m})$ is not too pessimistic of an upper bound when the pair of filtrations are truly random. In practice, when one has a time-varying filtration and the sampling points are relatively close [in time], the LCS between adjacent filtrations is expected to be much larger, subsequently making $d$ much smaller. 
For example, the average size of the LCS across the 10 filtrations from Section~\ref{sec:motivation} each having $m=417$ simplices sampled at evenly spaced time points was $343$, implying $d \approx 70$ permutations needed on average to update the decomposition between adjacent time points.

\subsection{Minimizing schedule cost}\label{sec:schedule_cost}
\subsubsection{Motivation}
The results from the previous section yield a sufficient algorithm for generating move schedules of minimal cardinality: simply compute a $\mathrm{LCS}(K_0, K_1)$ for a pair $(K_0, K_1)$ and fix an order of simplices in the set $\mathrm{LCS}(K_0, K_1) \smallsetminus K_0$ to move. Any sequence of moves which increases the LCS on each move is guaranteed to have size $m - \lvert \mathrm{LCS}(K_0, K_1) \rvert$, and the reduction to the permutation edit distance problem ensures this size is optimal. 
However, like in the vineyards algorithm, certain pairs of simplices cost more to exchange depending on whether they are critical pairs in the sense described~\cite{cohen2006vines}, making specific move operations more expensive than others. 
This small variation in cost-per-transposition induces a large variability in the cost of executing any given schedule, which is undesirable computationally. Ideally, we would like to generate a schedule which not only is minimal in size, but also cost. Formally, given two permutations $p, q \in S_m$, let $S = ( s_1, s_2, \dots, s_d )$ denote a sorting such that $S \circ p = q$. 
%The example given in section~\ref{sec:greedy} shows all (minimal size) $d!$ possible choices of $S$ for a fixed pair of permutations and choice of $\mathrm{LCS}(p,q)$. 
Let $\nu = \mathrm{LCS}(p,q)$ and let $\sigma \in S_d$ denote a choice of ordering for the symbols in  $p \smallsetminus \nu$, such that $S(\sigma)$ denotes a schedule generated from this ordering. Rephrasing from the description above, we would like to achieve something similar to the following objective: 
\begin{equation}\label{eq:schedule_opt}
	S_\sigma^\ast = \argmin_{\sigma \in S_d} \mathrm{cost}(S(\sigma) \circ p)
\end{equation}
where $\mathrm{cost}(S(\sigma) \circ p)$ measures the number of column operations require to execute $S(\sigma)$ on an existing decomposition. There are exponentially many ways to generate such a sorting $S$. 
% Dynamic programming discussion 
One advantage of Algorithm~\ref{alg:mr} is that the cost of applying a permutation to some given $R = D V$ decomposition can be determined efficiently prior to performing any column operations. 
Specifically, the cost of determining $\lvert \mathbb{I}\rvert + \lvert \mathbb{J}\rvert$ in lines (2) and (3) of Algorithm~\ref{alg:mr} cannot take more than $O(m)$ time. If we assume a similar complexity to determine the cost of applying a permutation with Algorithm~\ref{alg:ml} and we know to execute $d$ moves, one might expect be to use an optimization method like dynamic programming. 
That is, suppose $O(dm)$ move costs are computed ahead of time and these costs are stored in a table providing $O(1)$ element access and modification. Using a sparse row representation for $V$, this cannot take more than $O(dm)$ time, which is relatively efficient when $d << m$. With this structure, determining the cost of the cheapest move takes just $O(d)$ time. However, moving a simplex from position $i$ to position $j$ may require updating $O(\kappa^2)$ entries in the structure, where $\kappa = \lvert i - j \rvert$. Since $\kappa \sim O(m)$ in the worst case, direct optimization may require upwards of $\approx O(dm^2)$. Since this time complexity is potentially more expensive than executing a schedule to begin with, this direct strategy does not seem viable for practical settings.

@@

@@

When the set of permutations $S$ are adjacent transpositions, equality is achieved above since $K_\tau(\hat{s}_i, \hat{s}_{i+1}) = 1$ for every inversion between $p$ and $q$, though this not guaranteed if $S$ was derived using the strategy from section~\ref{sec:move_schedules}. Obtaining $S_\sigma^\ast$ via direct optimization over $S_d$ isn't practical, but experimental tests suggest that a greedy-type procedure performs well as a heuristic, which yields a $O(d^2 m \log m)$ minimization procedure, which is efficient when $d <<  m$.

@@

Although very similar to the Kendall-$\tau$ distance, the Spearman distance offers some convenient computational advantages: it can be computed in $O(m)$ time, the variance of its asymptotic distribution is larger than Kendalls for uniformly random permutations, and it satisfies a certain monotonicity property under LCS edit operations. 
Specifically, given two permutations $p,q \in S_m$ with $\nu = \mathrm{LCS}(p,q)$ between them, the analysis in section~\ref{sec:move_schedules} shows we must move each symbol $x \in \mathcal{D}$ where $\mathcal{D} = p \smallsetminus \nu $ in a way that extends $\nu$ after each move. Each symbol $x \in \mathcal{D}$ induces a permutation $s_x$ such that $F(p, q) \geq F(s_x \circ p, q)$. That is, unlike the Kendall distance, the Spearman distance satisfies: 
For any sorting $S$ such that $s_{i}$ that increases the size of longest common subsequence between $s_{i-1} \circ \dots \circ s_1 \circ p$ and $q$. Since the complexity of the move operations discussed in section~\ref{sec:moves} depend on the size of the movement, it is advantageous to construct a schedule which minimizes this movement. 
%Since $\lvert \mathcal{D} \rvert$ is fixed, we would like to choose $x \in \mathcal{D}$ such that the corresponding permutation $s_x$ reduces the total displacement of every symbol $x \in \mathcal{D}$ relative to  $q$. 
We consider the following objective modified from equation~\ref{eq:schedule_opt}: 

@@

The minimizer of this optimization can be interpreted as a sorting which minimizes the net displacement amongst all possible sortings, in the $\ell_1$ sense. This minimization problem is related to a constrained version of the bipartite crossing minimization problem for $k$ sets of permutations, which is NP-hard for $k \geq 4$~\cite{biedl2009complexity}. As a result, we instead consider a greedy approach. A greedy strategy requires $O((d+(d-1)+(d-2)+\dots+1)m) \approx O(d^2 m)$ time to complete, since there are $O(d)$ symbols to choose from and minimizing $F(\hat{s}_i, \hat{s}_{i+1})$ at each step takes $O(md)$. Since $\mathcal{D}$ is fixed, a greedy strategy thus depends on our ability to quickly compute $F(\hat{s}_i, \hat{s}_{i+1})$.  

@@

>>>>

% REQUIRED - "Abstracts must be able to stand alone and so cannot contain citations to the paper's references, equations, etc. 
\begin{abstract}
	The standard procedure for computing the persistent homology of a filtered simplicial complex is the matrix reduction algorithm. Its output is a particular decomposition of the total boundary matrix, from which the persistence diagrams and generating cycles can be derived. 
	Persistence diagrams are known to vary continuously with respect to their input, which motivates the algorithmic study of persistence computations for time-varying filtered complexes. Computationally, simulating persistence dynamically can be reduced to maintaining a valid decomposition under adjacent transpositions in the filtration order. In practice, the quadratic scaling in the number of  transpositions often makes this maintenance procedure slower than simply computing the decomposition from scratch, effectively limiting the application of dynamic persistence to relatively small data sets. In this work, we propose a coarser strategy for maintaining the decomposition over a discrete 1-parameter family of filtrations. Our first result is an analysis of a simple linear-time strategy which reduces the number of column operations needed to simulate persistence across a fixed homotopy by at most a factor of 2. We then show a modification of this technique which maintains only a sublinear number of valid states, as opposed to a quadratic number of states, and we provide tight lower bounds for this technique.
	Finally, we provide empirical results suggesting that the decrease in operations needed to compute diagrams across a family of filtrations is proportional to the difference between the expected quadratic number of states, and the proposed sublinear coarsening.
	Applications to multi-dimenisonal persistence and crocker stacks are also presented. 
\end{abstract}

@@

\section{Introduction} 
\subsection{Overview}\label{sec:overview} 
Given a triangulable topological space equipped with a sufficiently tame continuous function, persistent homology captures the changes in topology across the sublevel sets of the space, and encodes them in a persistence diagram. The stability of persistence contends that small changes in the underlying function produce small changes in the corresponding diagrams:  if the function changes continuously, so too will the points on the persistence diagram~\cite{cohen2007stability, cohen2006vines}. 
This motivates the study and application of persistence to time-varying or dynamic settings, like that of dynamic metric spaces~\cite{kim2020spatiotemporal}. 
As persistence-related computations tend to exhibit high algorithmic complexity (e.g. essentially cubic\footnote{For finite fields, it is known that the persistence computation reduces to the PLU factorization problem, which takes $O(m^\omega)$ where $\omega \approx 2.373$ is the matrix multiplication constant.} in the size of the underlying filtration~\cite{morozov2005persistence}), their adoption to dynamic settings poses a challenging computational problem.
With the current state of the art tools, there is no recourse to the problem of computing persistence from a time-varying filtration containing millions of simplices at thousands of snapshots in time.
An efficient means of updating the persistence of a [time-varying] filtration has far-reaching consequences: methods that vectorize persistence diagrams for machine learning purposes, like adaptive template functions~\cite{polanco2019adaptive} and persistence images~\cite{adams2017persistence}, or summary statistics like $\alpha$-smoothed Betti curves~\cite{ulmer2019topological}, all immediately become computationally viable tools in dynamic  settings. 
%This analogy provides a strong interpretation at gaining insight on the subtle topological changes that may occur in studying a continuous process.

Cohen-Steiner et al. refer to a continuous 1-parameter family of persistence diagrams as a \emph{vineyard}, and they give in \cite{cohen2006vines} an efficient algorithm for their computation given a time-varying filtration. 
The vineyards approach can be interpreted as an extension of the \emph{matrix reduction} algorithm~\cite{zomorodian2005computing}, which computes the persistence diagram of a fixed filtration $K$ with $m$ simplices in $O(m^3)$ time.
The main step in the reduction algorithm is a specific decomposition $R = D V$ of the boundary matrix $D$ of $K$. 
Since $V$ is always full rank (and upper-triangular), an alternative is to compute a $D = RU$ decomposition, where $U = V^{-1}$. 
The vineyards algorithm transforms a time-varying filtration into an ordered set of adjacent transpositions to be applied to the rows and columns of the decomposition $D = RU$, each of which take at most $O(m)$ time to execute. If one is interested in understanding how the persistent homology of a continuous function changes over time, then this algorithm is sufficient, for  homological critical points can only occur when the filtration order changes. 
This algorithm is also efficient  asymptotically: if there are $d$ time-points where this filtration order changes, then the vineyards algorithm takes $O(m^3 + md)$ time: one $O(m^3)$-time reduction upfront for the filtration at time $t_0$ followed by one $O(m)$ operation to update the decomposition for the remaining time points $(t_1, t_2, \dots, t_d)$. When $d >> m$, this $O(md)$ approach is far more efficient than the $O(dm^3)$  ``naive'' strategy of computing the diagrams independently at every time point.

Despite its theoretical efficiency, vineyards is often not the method of choice in practical settings. 
Indeed, while there is an increasingly rich ecosystem of software packages offering variations of the standard reduction algorithm (e.g. Ripser, PHAT, Dionysus, etc... see~\cite{otter2017roadmap} for an overview), implementations of the vineyards algorithm are relatively uncommon.\footnote{Dionysus 1 has an implementation of vineyards, but the algorithm was never ported to the 2nd version.} The reason for this disparity is perhaps explained by Lesnick and Wright~\cite{lesnick2015interactive}: ``While an update to an $RU$ decomposition involving few transpositions is very fast in practice... many transpositions can be quite slow... it is sometimes much faster to simply recompute the $RU$-decomposition from scratch using the standard persistence algorithm.'' Indeed, whether performing the reductions naively or updating the decomposition dynamically using vineyards, they observed that obtaining the decomposition is the most computationally demanding aspect of their 2-D persistence algorithm.

This work seeks to further understand and remedy this discrepancy: building on the work presented in~\cite{busaryev2010tracking}, we introduce a coarser approach to the vineyards algorithm.
While vineyards are  efficient for   \emph{continuous} time families of diagrams, it is not necessarily efficient when the time parameter is coarsely discretized.
Our methodology is based on the observation that practitioners often don't need (or want!) all $d$ persistence diagrams generated by a continuous 1-parameter of filtrations; usually just $n << d$ of them are sufficient.   
% in many time varying settings, a valid decomposition is often only sought after at a relatively sparse number of time points relative to the size of the (finite) set needed to generated by the family. 
By exploiting the ``donor'' concept introduced in in~\cite{busaryev2010tracking}, we are able to make a tradeoff in the number of times the decomposition is restored to a valid state, dramatically reducing the total number of column operations needed to apply an arbitrary permutation to the filtration. This trade off, paired with a fast greedy heuristic explained in section~\ref{sec:proxy_objective}, yields an algorithm that can update a $R = DV$ decomposition more efficiently than vineyards in coarse contexts, making dynamic persistence more computationally tractable for a wider class of use-cases.

@@

 \subsection{A Motivating Example and Contributions}\label{sec:motivation} 
As a motivating example, we describe a simple experiment illustrating why the vineyards algorithm does not always yield an efficient method for time-varying applications. 
Consider a series of grayscale images (i.e. a video) depicting a fixed-width annulus expanding about the center of a $9 \times 9$ grid, where scale of the annulus is parameterized by a single parameter $\tau$. Thus, the data set is a 1-parameter family of images. Each image contains the same number of pixels whose intensities---these are determined by $\tau$---vary with time. For each image, we build a filtered simplicial complex using the \emph{Freudenthal} triangulation, with the ordering of inclusions obtained by lower stars.  
 \begin{figure*}[t]
 	\centering
 	\includegraphics[width=0.36\textwidth, height = 1.45in]{circle_vineyards.png}
 	\includegraphics[width=0.63\textwidth, height = 1.45in]{circle_complex.png}
 	\caption{On the right, the grayscale image data set and corresponding lower star filtrations. On the left, the cumulative number of column operations required to compute persistence for this time-varying filtration. The persistent Betti numbers for each filtration from left to right are: $\beta_1 = (1,0,0), \beta_2 = (1,1,0), \beta_3 = (1,1,0), \beta_4 = (1,1,0), \beta_5 = (4,0,0)$. Observe computing $10$ diagrams independently at evenly spaced time points (the green line) both captures these major topological changes and is the most computationally efficient approach shown.}
 	\label{fig:vineyards}
 \end{figure*}
Snapshots of the image data and its corresponding time-varying filtration is depicted in Figure~\ref{fig:vineyards} (top right and bottom right, respectively), which we use as a baseline for comparing the standard algorithm $\texttt{pHcol}$ with vineyards. Note the size of the filtration is fixed--no simplices are ever added or removed over time. 
There are two events which dramatically alter the persistence diagrams extracted from these filtrations: the first occurs when the central connected component splits to form a cycle, and the second occurs when the annulus approaches the outer dimensions of the grid, splitting into four components. As a consequence, in this example, only a few persistence diagrams are needed to capture the major topological changes over time.

  Consider the situation where one is concerned with capturing these dramatic changes in homology.
  On the left side of Figure~\ref{fig:vineyards}, we compare the cumulative cost (in total number of column operations) of various approaches which one can use to simulate persistence dynamically. 
  Since it is unknown a priori at which times  the persistent homology will change, one solution is to discretize the time domain at $n$ evenly spaced points and compute the persistence diagrams at each one of them  independently. An alternative approach is to construct a homotopy between two time-points, and then decompose this homotopy into adjacent transpositions--the vineyards approach. The former is often used in practice, but it is only an approximation; the latter is guaranteed to capture all homological changes in persistence that occur in simulating the 1-parameter family. The graph in Figure~\ref{fig:vineyards} depicts two different size approximations (purple, $n=100$ and green, $n=10$), wherein the reduction algorithm (\texttt{pHcol}) was applied at $n$ time points, and two exact strategies (black and blue), which compute all $\approx 7 \times 10^{4}$ diagrams given by the homotopy. As one can see from the figure, if one needs to simulate the homotopy exactly, the vineyards approach is indeed far more efficient than naively applying the reduction algorithm independently at all time points.
  However, when the discretization of the time domain is sparse enough, the naive approach actually performs less column operations than the vineyards strategy, while still capturing the main events.

@@

The original purpose of the vineyards algorithm, as described in~\cite{cohen2006vines}, was to compute a continuous 1-parameter family of persistence diagrams given a time-varying filtration, and to detect homological critical events during the construction.

@@

\begin{corollary}\label{cor:expectation}
	If $K_0, K_1$ are random distinct filtrations each of size $m$, then $d$ is upper bounded by $O(m - \sqrt{m})$ in expectation, with probability $1$ as $m \to \infty$.
\end{corollary}
\begin{proof}
	The proof of this result reduces to showing the average length of the LIS for random permutations. Let $L(p) \in [1,m]$ denote the maximal length of a increasing subsequence of $p \in S_m$. 
	The essential quantity to show the expected length of $L(p)$ over all permutations: 
	$$ \ell_m = \mathbb{E} \, L(p) = \frac{1}{m!} \sum\limits_{p \in S_m} L(p)$$
	The history of the mathematics of estimating this quantity dates back at least 50 years, and is sometimes called the \emph{Ulam-Hammersley} problem. The seminal work by Baik et al.~\cite{baik1999distribution} established that as $m \to \infty$:
	$$ \displaystyle \ell_m = 2 \sqrt{m} + c m^{1/6} + o(m^{1/6}) $$
where $c = -1.77108...$. As a result, we have: 
$$ \frac{\ell_m}{\sqrt{m}} \to 2\rlap{\quad \text{as } $m \to \infty$} $$	 
Thus, if $p \in S_m$ denotes a uniformly random permutation in $S_m$, then $L(p)/\sqrt{m} \to 2$ in probability as $m \to \infty$. 
Since the schedule size is defined as $d = m - \lvert \mathrm{LCS}(K_0, K_1) \rvert$, the claimed bound follows.
\end{proof}
\noindent The upper bound from Corollary~\ref{cor:expectation} captures the size of $S^\ast$ under worst-case conditions. It is worth noting that this bound applies to pairs of uniformly sampled permutations, as opposed to uniformly sampled distinct filtrations---intuitively, simplicial filtrations have more structure due to the subcomplex relation that must be respected in the filtration. Thus, while $O(m - \sqrt{m})$ in expectation is an improvement over $O(m)$, it is still not tight. However, Boissonnat~\cite{boissonnat2018efficient} has shown that for any $d$-dimensional simplicial complex $K$ with $m$ simplices, the number of distinct filtrations built from $K$ is \emph{at least} $\floor{\frac{t+1}{d+1}}^m$, where $t$ is the number of distinct filtration values associated with the complex $K$. Since this lower bound grows similarly to $m!$ when $t \sim O(m)$ and $d << m$ fixed, we expect $O(m - \sqrt{m})$ is not too pessimistic of an upper bound when the pair of filtrations are truly random. In practice, when one has a time-varying filtration and the sampling points are relatively close [in time], the LCS between adjacent filtrations is expected to be much larger, subsequently making $d$ much smaller. 
For example, the average size of the LCS across the 10 filtrations from Section~\ref{sec:motivation} each having $m=417$ simplices sampled at evenly spaced time points was $343$, implying $d \approx 70$ permutations needed on average to update the decomposition between adjacent time points.

\subsection{Minimizing schedule cost}\label{sec:schedule_cost}
\subsubsection{Motivation}
The results from the previous section yield a sufficient algorithm for generating move schedules of minimal cardinality: simply compute a $\mathrm{LCS}(K_0, K_1)$ for a pair $(K_0, K_1)$ and fix an order of simplices in the set $\mathrm{LCS}(K_0, K_1) \smallsetminus K_0$ to move. Any sequence of moves which increases the LCS on each move is guaranteed to have size $m - \lvert \mathrm{LCS}(K_0, K_1) \rvert$, and the reduction to the permutation edit distance problem ensures this size is optimal. 
However, like in the vineyards algorithm, certain pairs of simplices cost more to exchange depending on whether they are critical pairs in the sense described~\cite{cohen2006vines}, making specific move operations more expensive than others. 
%This variation in cost exaggerates the variability of the cost of scheduling move operations.

One advantage of Algorithm~\ref{alg:mr} is that the cost of applying a permutation to some given $R = D V$ decomposition can be determined efficiently prior to performing any column operations. 
Specifically, the cost of determining $\lvert \mathbb{I}\rvert + \lvert \mathbb{J}\rvert$ in lines (2) and (3) of Algorithm~\ref{alg:mr} cannot take more than $O(m)$ time. If we assume a similar complexity to determine the cost of applying a permutation with Algorithm~\ref{alg:ml} and we know to execute $d$ moves, an initial thought might be to use an optimization method like dynamic programming. 
That is, suppose $O(dm)$ move costs are computed ahead of time and these costs are stored in a table providing $O(1)$ element access and modification. Using a sparse row representation for $V$, this cannot take more than $O(dm)$ time, which is relatively efficient when $d << m$. With this structure, determining the cost of the cheapest move takes just $O(d)$ time. However, moving a simplex from position $i$ to position $j$ may require updating $O(\kappa^2)$ entries in the structure, where $\kappa = \lvert i - j \rvert$. Since $\kappa \sim O(m)$ in the worst case, direct optimization may require upwards of $\approx O(dm^2)$. Since this time complexity is potentially more expensive than executing a schedule to begin with, this direct strategy does not seem viable for practical settings.

@@

As before, given two permutations $p, q \in S_m$, let $S = ( s_1, s_2, \dots, s_d )$ denote a sorting such that $S \circ p = q$. 
In general, there are exponentially many ways to generate such a sorting $S$. 
The example given in section~\ref{sec:greedy} shows all (minimal size) $d!$ possible choices of $S$ for a fixed pair of permutations and choice of $\mathrm{LCS}(p,q)$. 
Let $\tau = \mathrm{LCS}(p,q)$ and let $\sigma \in S_d$ denote a choice of ordering   for the symbols in  $p \smallsetminus \tau$, 
such that $S(\sigma)$ denotes a schedule generated from this ordering. Ideally, we would like to achieve something similar to the following objective: 
\begin{equation}\label{eq:schedule_opt}
	S_\sigma^\ast = \argmin_{\sigma \in S_d} \mathrm{cost}(S(\sigma) \circ p)
where $\mathrm{cost}(S(\sigma) \circ p)$ measures the number of column operations require to execute $S(\sigma)$ on an existing decomposition. 

@@

When the set of permutations $S$ are adjacent transpositions, equality is achieved above since $K_\tau(\hat{s}_i, \hat{s}_{i+1}) = 1$ for every inversion between $p$ and $q$. However, this is not guaranteed if $S$ was derived using the strategy from section~\ref{sec:move_schedules}. Obtaining $S_\sigma^\ast$ via direct optimization over $S_d$ isn't practical, but experimental tests suggest that a greedy-type procedure performs well as a heuristic, which yields a $O(d^2 m \log m)$ minimization procedure. When $d <<  m$, this strategy of successively minimizing the Kendall-$\tau$ distance is efficient. 

@@

Although very similarity to the Kendall-$\tau$ distance, the Spearman distance offers some convenient computational advantages: it can be computed in $O(m)$ time, and the variance of its asymptotic distribution is larger. Consider the following minimization: 

@@

The minimizer of this optimization can be interpreted as a sorting which minimizes the net displacement amongst all possible sortings, again in the $\ell_1$ sense. 

@@

Given two permutations $p,q \in S_m$ and a LCS $\tau = \mathrm{LCS}(p,q)$ between them, the analysis in section~\ref{sec:move_schedules} shows we must move each symbol $x \in \mathcal{D}$ where $\mathcal{D} = p \smallsetminus \tau$ in a way that extends $\tau$ after each move. Thus we have $d = \lvert \mathcal{D} \rvert$ symbols to choose from initially. Each symbol $x \in \mathcal{D}$ yields a permutation $s$ to apply to $p$. Unlike the Kendall distance, the Spearman distance satisfies: 
